### 2009-09-01 trunk/latex/paper1/contents.tex
incluindo esboco de citacoes e figuras.

Software modularization recovery algorithms automatically recognize a system's
modular structure by analyzing its implementation.
Due to the lack of well document software systems, though, the issue of testing
these algorithms is still underexplored, limiting both their adoption in the
industry and the development of better algorithms.
We propose to rely on software models to produce arbitrarily large test sets. In
this paper we consider three such models and analyze how similar the artifacts
they produce are from artifacts from real software systems.

@@

Unfortunately, in the case of legacy systems, such knowledge isn't available.
Depending on its size, it might take months to understand the system so well as
to find a good modularization.

@@

The empirical evaluations consist of selecting a collection of systems with
known reference modularizations and then applying the algorithms to the systems.
The modularizations found by the algorithms are then compared to the reference
decompositions by a metric such as MoJo CITE or PrecisionRecall CITE.

@@

\section{Software Networks and Software Clustering}

@@

\section{Complex Networks}

Complex network theory found many scale-free networks

Software dependency networks are scale free. CITE Valverde, Myers

Scale free means ... $N(k) \sim k^{-\gamma}$

\section{Models}

Many scale-free models have been proposed. Only a few, though, produce modular
networks.

@@

@@

* With probability p, a new vertex is added to the network, together with an
edge from the new vertex to an existing vertex, chosen preferentially according
to .... Considering din = 0, this means that a vertex with indegree = 8 is
twice more likely to receive the edge than a vertex with indegree = 4. The
parameter din can be used to alleviate the handicap. (If din = 4, the first
vertex will only be 3/2 more likely to receive the edge). The new vertex is
put on the same cluster as the old vertex.
* With probability q, a new vertex is added to the network, together with an
edge from an existing vertex, chosen preferentially accordingly to dout +
outdegree, to the new vertex. The new vertex is put on the same cluster as the
old vertex. This case is similar to the previous case.
* With probability r, a new edge is added between two vertices, v and w. v is
chosen according to .... After that, with prob X, w is chosen from one of the
modules connected to v's module, else it's chosen from v's own module. In any
case, the exact choice is made according to ...

@@

We want to investigate if the models can produce networks that resemble software
networks. We know that they share with software networks the scale-free
property. This is not enough, since many real networks share this property. So
we looked for a method to differentiate between software networks and other
networks.

What we are looking for, after all, is an oracle that accepts software networks
and rejects non-software networks. If the oracle has these two properties, we
can be confident that it'll accept only synthetic networks that resemble
software networks.

In a recent work, Milo et al. proposed the study of triads in order to
characterize different classes of networks. We thus follow their work here.

Triads are...

Figure 1a show triads for a software system... Figure 2a show triads for network
from domain X.

\subsection{Software Network Extraction}

We've collected 65 systems written in Java and X networks from many domains,
such as sociology, biology, technology and linguistics.
Table 1 ...

\subsection{Oracle Training and Evaluation}

We then used Pearson's correlation coefficient as a similarity measure between
two networks. For each software network, We then computed the average
correlation coefficient (ACC) to the other software networks. We've observed
that among software networks the ACC is X +- Y.

We then computed, for each non-software network, the ACC to the 65 software
networks. By the 3-sigma rule, we use X as the threshold for realistic software
networks: networks whose ACC is below this threshold are rejected.

The oracle has X precision and X recall...

@@

We chose five different module dependency networks, which where extracted from
actual dependencies between modules of five different software systems of our
sample, ranging from 2 to 32 modules. The module dependency networks are shown
on Figure XXX. 
%02-gef
%04-ibatis
%08-megamek
%16-findbugs
%32-zk

@@

Total: 1296 networks

@@

p1,p2,p3,p4 ranging from 0.0 to 1.0 in 0.2 steps, with p1 > 0, p1 + p2 + p3 + p4 = 1.0.
e1,e2,e3,e4 in 1, 2, 4, 8 (except that when pi = 0, ei is ignored).
2*p4*e4 >= p1*e1 + p2*e2, so the number of edges created is at least the 
double of the number of edges removed
alpha in -1, 0, 1, 10, 100, 1000. 
m in 2, 4, 8, 16, 32 (just like bcr and lfr)

Total: 38790 networks

\subsubsection{ER}

Number of edges chosen randomly in the range ...

\section{Realism Evaluation}

@@

Show some graphs: histogram of average correlations for each model.

As expected, ER produces non-realistic software networks.

@@

The use of models lead to a more controlled way of experimented, and allows us
to play with a large data set and a wide variety of characteristics. Although
we have shown that the models studied can produce realistic networks according
to one criterium, we can't just discard the experimentation with real software
systems, as they are the subjects in which algorithms will be used in the
real world. In particular, it's an open question whether the models can 
reproduce any software system or if they just generate a subset of systems.
We believe that there are software systems that cannot be simulated by none
of these models.

Although common in distributed computing research, simulation is underexplored
in software engineering. This work opens a door to the usage of simulation in
the field of reverse engineering of software in order to evaluate RE 
algorithms.

Future work: to apply algorithms to the networks and compare the results with
results found in the literature.

>>>>

Software modularization recovery algorithms automatically recognize a system's
modular structure by analyzing its implementation.
Due to the lack of well document software systems, though, the issue of testing
these algorithms is still underexplored, limiting both their adoption in the
industry and the development of better algorithms.
We propose to rely on software models to produce arbitrarily large test sets. In
this paper we consider three such models and analyze how similar the artifacts
they produce are from artifacts from real software systems.
\cite{Pollet2007}

@@

Unfortunately, in the case of legacy systems, such knowledge isn't available.
Depending on its size, it might take months to understand the system so well as
to find a good modularization. XXX \cite{Parnas1972}

@@

The empirical evaluations consist of selecting a collection of systems with
known reference modularizations and then applying the algorithms to the systems.
The modularizations found by the algorithms are then compared to the reference
decompositions by a metric such as MoJo \cite{Tzerpos1999} or PrecisionRecall.

% TODO: diagram showing the evaluation of a clustering algorithm

@@

\section{Software Modularization Recovery}

Aka software architecture recovery, software architecture reconstruction,
software clustering...

It's an task of reverse engineering, as defined by Tonella \cite{Tonella2007}...

%\section{Software Networks and Software Clustering}

@@

\section{Networks Theory}

Network theory research studies general properties of many types of networks by
using statistical analysis. In the last decade, it has been found that many
networks arising from sociology, biology, technology and other domains have
remarkable similarities. It has been shown that, in these networks, the number
of vertices connected to k edges, N(k), is proportional to $k^{\-gamma}$, where
$\gamma$ is a positive constant. These networks were called scale-free networks.

Network theory has been applied to software networks and it was shown that they
are also scale-free networks \cite{Myers2003,Valverde2003}.

(in fact we can't argue the degree distribution is perfectly fit by a power law.
Anyway, the distribution is much more assymetric than normal ou Poisson's)

software indegree distribution (power law), outdegree distribution (not so power
law)

\section{Network Models}

Many models were proposed to explain the formation of scale-free networks. These
models are simple algorithms that can be proven, either formally or emprically,
to generate networks that are scale-free.

@@

out-neighbors(v) 
module(v)
neighbor-modules(v)

choose event with probabilities (p, q, r)
event 1:
  choose w within V according to f(x) = din + in-degree(x)
  add new vertex v to the module of w
  add edge from v to w
event 2:
  choose w within V according to f(x) = dout + out-degree(x)
  add new vertex v to the module of w
  add edge from w to v
event 3:
  choose v within V according to f(x) = dout + out-degree(x)
  choose case with probabilities (mu, 1 - mu)
  case 1: % out
    choose w within neighbor-modules(v) - neighbors(v) according to f(x) = din + in-degree(x)
    add edge from v to w
  case 2:
    choose w within module(v) - neighbors(v) according to f(x) = din + in-degree(x)
    add edge from v to w
end

@@

* Add a node with ongoing edge. With probability p, a new vertex is added to the network, together with an
edge from the new vertex to an existing vertex, chosen preferentially according
to .... Considering din = 0, this means that a vertex with indegree = 8 is
twice more likely to receive the edge than a vertex with indegree = 4. The
parameter din can be used to alleviate the handicap. (If din = 4, the first
vertex will only be 3/2 more likely to receive the edge). The new vertex is
put on the same cluster as the old vertex.

* Add a node with ingoing edge. With probability q, a new vertex is added to the network, together with an
edge from an existing vertex, chosen preferentially accordingly to dout +
outdegree, to the new vertex. The new vertex is put on the same cluster as the
old vertex. This case is similar to the previous case.

* Add an edge. With probability r, a new edge is added between two vertices, v and w. v is
chosen according to .... After that, with prob X, w is chosen from one of the
modules connected to v's module, else it's chosen from v's own module. In any
case, the exact choice is made according to ...

@@

% Is there such a thing as softwareness? Can we reproduce it?

``softwareness''

Our hypothesis is that at least one of the presented models can synthesize
networks that resemble software networks. We already know that they produce
networks that, just like software networks, are scale-free. This single
property, though, isn't enough to prove the hypothesis, since there are many
known scale-free networks extracted from many distinct domains of knowledge. It
is important to distinguish between software and non-software networks.

% v.  differentiate, differ; distinguish; contrast; discern

The approach we chose to test our hypothesis was to build a network classifier:
an algorithm that analyses a network and classifies it as a software network or
as a non-software network. If a classifier is capable of classifying with high
accuracy both software and non-software networks, then we can apply it to
synthetic networks and see if they are discernable from software networks.
it gives a fair approximation apply it on synthetic networks. We expect that
some synthetic networks will be so similar to software networks that they will
be classified as software networks.

The experiment we conducted to evaluate the models, described in details in the
following sections, can be summarized as such:
* First of all, we collected both software and non-software networks
* Then, we devised a classifier
* After that, we applied the classifier to our data set and evaluated its
accuracy, precision and recall.
* After we came to an acceptable classifier, we synthesized many networks by
varying the parameters on the models we presented.
* Finally, we classified all synthesized networks as software or non-software.

After that we analysed the results and tried to understand which parameters
of each model contribute mostly to the softwareness of their networks

%First of all, we need a data set containg both software and non-software
%networks. 
%Therefore, in order to prove our hypothesis we must show that, according to some
%criterion, some of at least some of
%the synthetic networks are more similar to software networks than other networks
%are.
%What we are looking for, after all, is an oracle that accepts software networks
%and rejects non-software networks. If the oracle has these two properties, we
%can be confident that it'll accept only synthetic networks that resemble
%software networks.

\subsection{Real Networks Data Set}

For this experiment we have collected 65 software systems written in the Java
programming language and 66 networks from other domains, such as biology,
sociology and linguistics. Both the software systems and the networks are freely
available on the Internet.

As of this writing, Java is a popular programming language, for which there is
plenty of open source software CITE. Also, there are many dependency extractors
for this language. We have extracted the dependency network for each software
system using a tool called Dependency Finder CITE, for its easy integration with
scripts. Dependency Finder looks for dependencies by means of static analysis of
Java bytecode and outputs the dependency network in a XML file. Although this
network is in a detailed level, showing dependencies between fine-grained
implementation entities such as attributes and methods, we have abstracted it to
a network of dependencies between classes. The network is directed and
non-weighted. The size of the networks range from XXX to YYY.

% JARs?

The networks from other domains were collected in websites of complex network
research groups. The domains include sociology, linguistics, biology and
technology. The size of the networks range from XXX to YYY vertices.

\subsection{Classifier} % Software network classifier

In a recent work, Milo et al. proposed to distinguish networks of different
domains by analyzing the triad concentration of each network. A triad is simply
a network with three vertices in which all vertices are connected. There are
only 13 distinct triads -- one for each configuration of directed edges --, as
shown in Figure \ref{fig:triads}.
\cite{Milo2002}
%CONSIDERING all the possible configurations of
%directed edges between three vertices, there are to 13 triads in which all
%vertices are connected, as shown in Figure X.

\begin{figure*}[!t]
\centering
\includegraphics[width=1.0\textwidth]{triads}
\caption{Triads (id vs. graph. reprs.)}
\label{fig:triads}
\end{figure*}

In order to characterize a network, one can count how many times each of the
13 triads appear on the network. The result is a triad concentration profile.
Figure 1a show triads for a software system... Figure 2a show triads for network
from domain X. \ref{fig:profiles}

\begin{figure*}[!t]
\center
\subfigure[ref1][Software
network]{\includegraphics[width=2.5in]{triads-software}}
\hfil
\subfigure[ref2][Non-software network]{\includegraphics[width=2.5in]{triads-nonsw}%
}
\caption{Triad profiles}
\label{fig:profiles}
\end{figure*}

Then, in order to measure how similar two networks are, Milo et al computed
Pearson's correlation coefficient between the two corresponding profiles.

CITE igraph

In order two measure the softwareness of a network, we measure the correlation
between the network's profile and the profile of each of the 65 software
networks in our sample. We then compute the average of these correlations and
call it the S-score, that varies between 0 and 1. The greater the S-score, more
likely that the network is a software network.

In order to build our classifier, we need a threshold for the S-score, so
networks are classified as software networks only when they have S-score greater
or equal to this threshold. We computed the S-score for each of the 65 software
systems and found an average S-score of 0.97 with standard deviation of 0.03.
Using the three-sigma rule from statistics, we chose the threshold as 0.97 - 3 *
0.3, that is 0.88.

The small standard deviation shows that triad concentration profiles indeed
characterize well the software domain.

\subsubsection{Classifier Validation}

In order to assess if the classifier is capable of distinguishing between
software and non-software networks, we applied it to all the networks from our
sample.

Most software networks were classified correctly, which is not surprising since
the classifier was based on them.

Precision: X, recall: Y

Curiously, polblog had a very high S-score. This issue requires further 
investigation. Anyway, we cannot deny the possibility that it is really very
software-like.

We could have chosen another value for the threshold, and it would affect the
precision and the recall. For instance, had we chosen 0.82, we would end with
100\% recall and x\% precision. For us, though, it is more important to have a
high precision, so we keep the threshold 0.88.

We also generated networks with the ER model, which is known to generate
networks that are not scale-free and, thus, are very different from software
networks. All the ER networks were classified as non-software.

%Following Milo et al., We then used Pearson's correlation coefficient as a
%similarity measure between two networks. For each software network, we then
%computed the average correlation coefficient (ACC) to the other software
%networks. We've observed that among software networks the ACC is X +- Y.

% avg: 0.97. stdev: 0.03
%We then computed, for each non-software network, the ACC to the 65 software
%networks. By the 3-sigma rule, we use X as the threshold for realistic software
%networks: networks whose ACC is below this threshold are rejected.
%The oracle has X precision and X recall...

@@

We chose five different module dependency networks, which where extracted from
actual dependencies between modules of five different software systems of our
sample, ranging from 2 to 32 modules. The module dependency networks are shown
on Figure \ref{fig:architectures}.

\begin{figure*}[!t]
\centering
\includegraphics[width=1.0\textwidth]{architectures}
\caption{Module dependency graphs}
\label{fig:architectures}
\end{figure*}

%02-gef
%04-ibatis
%08-megamek
%16-findbugs
%32-zk

@@

Total: 1,296 networks

@@

p1,p2,p3,p4 ranging from 0.0 to 1.0 in 0.2 steps, with p1 > 0, p1 + p2 + p3 + p4 = 1.0.
e1,e2,e3,e4 in 1, 2, 4, 8 (except that when pi = 0, ei is ignored).
2*p4*e4 >= p1*e1 + p2*e2, so the number of edges created is at least twice
the number of edges removed (to avoid long run times)
alpha in -1, 0, 1, 10, 100, 1000. 
m in 2, 4, 8, 16, 32 (just like bcr and lfr)

Total: 38,790 networks

\section{Softwareness Evaluation}

We used our classifier ...

@@

Show some graphs: histogram of average correlations for each model.
\ref{fig:histograms}

\begin{figure*}[!t]
\center

\subfigure[Software network]{\includegraphics[width=1.0in]{hist-cgw}}

\subfigure[Non-software network]{\includegraphics[width=1.0in]{hist-bcr}}

\subfigure[Non-software network]{\includegraphics[width=1.0in]{hist-lf}}

\caption{Triad profiles}
\label{fig:histograms}
\end{figure*}

@@

We have shown empirically that network models found in the literature can
produce synthetic networks that are similar to the network of dependencies
between classes in object-oriented systems. This result supports the use of
synthetic networks in the evaluation of software modularization algorithms.
%that operate on class dependency networks.

% review
The use of synthetic data is common in distributed computing research, but still
underexplored in software engineering research. Since many reverse engineering
tasks extract from component dependencies information useful for the maintenance
of a software system, we expect this work to have impact beyond the software
modularization recovery community.

%Although the use of models and synthetic data is somewhat common in distributed
%computing research, it is underexplored in software engineering research. We
%expect this work to contribute to explore t
%common in distributed computing research, simulation is underexplored
%in software engineering. This work opens a door to the usage of simulation in
%the field of reverse engineering of software in order to evaluate RE 
%algorithms.

We accept that it is very important to test the algorithms with networks
extracted from real software systems, but we argue that the use of synthetic
networks in a complementary manner can give researchers new insights about the
algorithms. First, the use of models allows the creation of large test sets,
thus diminishing the small sample effects. Moreover, the networks are created in
a controlled way, accordingly to model parameters, so it is possible to study
the behavior of the algorithms with different parameter values.

In a future work, we intend to evaluate with synthetic networks some software
clustering algorithms that were previously studied with real data. We will,
then, be able to compare the results obtained by both approaches. \cite{Wu2005}
